MetaTracts - A Method for Robust Extraction and Visualization of Carbon Fiber Bundles in Fiber Reinforced Composites

### 2014-09-26 chapters/MetaTracts.tex
supplemental folders

Fig.~\ref{fig:data-char} provides an example of our targeted dataset. It also shows the recurring fiber bundle pattern in the final CFRP laminate, called a \textit{unit cell}.

@@

\item{ How to visualize the interaction between a particular pair of fiber bundles (crossovers/braiding) or a unit cell?}

@@

Answering the above queries from the volume renderings of XCT datasets, or from visual inspection of particular 2D slices is non-trivial, even for experts. We detail and evaluate a method which uses visualization techniques to gain insight into our data. We demarcate the above queries into two parts: \textit{geometric structure} and \textit{spatial context}. Geometric structure refers to the shape, size, and orientation of a fiber bundle or a group of bundles. Spatial context refers to how two or more bundles interact with each other. 
We advance and interpret techniques from diffusion tensor imaging to extract and visualize geometric structures from 3D X-ray computed tomography data of the woven carbon fiber reinforced composites. The main goal of this work is to expand the state of the art in non-destructive testing through visualization of composite structures in \textit{lower resolutions} and thus \textit{larger areas} of woven fabrics.
\begin{figure}[tb]
	\centering
	%\includegraphics[width=0.45\textwidth]{imagesMT2014/algo.png}
	%\includegraphics[width=0.5\textwidth,  trim = 0mm 300mm 0mm 0mm, clip,]{imagesMT2014/image_algo}
	\includegraphics[width=\linewidth]{images_pvis/workflow.pdf}
	\caption{Flow chart of the MetaTracts approach for fiber bundle extraction}
	\label{fig:flowchart}
\end{figure}
Fig.~\ref{fig:flowchart} shows an over view of our approach. Starting from the XCT data, we first generate MetaTracts, cluster the MetaTracts to create fiber bundles and then generate visualizations from the clusters. For a detailed description of the MetaTracts generation process, see Sec.~\ref{subsec:MetaTracts-generation}. For the description of plausible fiber bundle generation see Sec.~\ref{subsec:fiber-bundles}. For a discussion of the visualizations we create for our domain experts see Sec.~\ref{sec:vis}. Sec.~\ref{sec:results} describes our experimental results, Sec.~\ref{sec:param_choices} presents our parameter choices and  Sec.~\ref{sec:user_eval} summarizes our user evaluations.

@@

\subsection {Visual Analysis of Fiber Reinforced Composites}
The approaches presented in visualization and analysis of composites mainly focus on individual objects such as fibers or pores and deals with fiber extraction from high resolution data where the individual fibers are clearly discernible. Fritz et al.~\cite{Fritz2009} proposed interactive workflows for non destructive testing practitioners to explore and quantify steel fibers in reinforced sprayed concrete. This approach allows analyzing fiber orientations based on direction transfer functions. Salaberger et al.~\cite{Salaberger2011} introduced a pipeline to extract and characterize individual fibers of fiber reinforced composites. They encode the extracted fibers as color-coded line segments in 3D and visually identifying fibers with similar orientations. Reh et al.~\cite{PMI_AR_2012} introduced an approach to explore pores of carbon fiber reinforced composites. Recently, Weissenbock et al.~\cite{Weissenbock2014} introduced a system for interactive exploration and analysis of fibers in fiber reinforced polymers. 
Tools based on image analysis, fiber tracking and clustering into bundles in DTI and their subsequent visualization through stream tubes and coloring schemes have received extensive research. Recently, similar techniques have been applied to fiber tracing in composite materials. To the best of our knowledge there is no approach focusing on direct extraction of fiber bundle structures from low resolution XCT of composite fibers, which is the main scope of this work.

@@

\begin{figure}[tb]
\centering
%\includegraphics[width=0.5\textwidth, trim = 0mm 110mm 0mm 0mm, clip,]{images_pvis/figure1}
\includegraphics[width=\linewidth]{images_pvis/data-char.pdf}
\caption{Data characteristics: (a) Rendering of the data. (b) 2D slices along the X-axis and (c) 2D slice along the Z-axis. (d) Magnified region marked in green square (c). Multiple fiber bundles cross and are indistinguishable by visual inspection alone. }
\label{fig:data-char}
\end{figure}
\section {Data Characteristic and Assumptions}
\label{sec:char_data}
Fig. \ref{fig:data-char} shows dataset 1 with woven fiber bundles. Dataset 1 is 450$\times$300$\times$500 voxels in size with isotropic resolution of 2$\mu m$. The data type is unsigned 8 bit integer. Fig.~\ref{fig:data-char} clearly shows the weaving structure of the composite unit cell as regional subset of the recurring fiber bundle weaving pattern. The unit cell is used as a basis structure for CFRP manufacturing. The alignment and weaving pattern of the fiber bundles in the cloth influences the strength and stiffness properties of the resulting material. Fig.~\ref{fig:data-char}a shows a volume rendering of the dataset. Fig.~\ref{fig:data-char}b shows 2D slices along the $X$- and Fig.~\ref{fig:data-char}c in $Z$-axis. Fig.~\ref{fig:data-char}d shows the magnified version of the green region of interest. 
The resolution of dataset 1 renders the individual carbon fibers in a fiber bundle hard to resolve in the 2D slice images. Fig.~\ref{fig:data-char}d contains two bundles going in opposite directions. The separation between two fiber bundles is barely visible. The carbon fibers themselves do not differ much from the underlying epoxy matrix in terms of attenuation which poses an additional challenge for characterization. The fiber bundles may differ in terms of the amount of fibers in the bundle. Fig.~\ref{fig:data-char}a shows the large variation in cross section sizes among the bundles.
Depending on the weaving pattern, the fiber bundles cross each other in different orientations. In general, the number of orientations is limited by the weaving pattern. Nevertheless, the weaving pattern may cause the individual bundles to be curved. In consequence individual fibers may be adjacent in Euclidean distance but belong to different bundles. 
We make the following assumptions on our data:
\begin{itemize}[noitemsep]
\item The structure embedded in the data contains fiber bundles of indiscernible fibers.
\item Local orientation: each point within a fiber bundle has a local orientation which is parallel to the corresponding area in the fiber bundle.
\item The local orientation may change gradually inside the fiber bundle.
\item The local orientation may be noisy and not reliable.
\item Connectivity: moving along the direction of a non-noisy local orientation in small increments, we will reach another neighborhood with similar local orientation.
\item Fiber bundles going in different directions only interact near the surface of contact.
\end{itemize}
%\begin{figure}[htp]
%  \centering
%    \subfloat[Overview of the dataset]{\label{figur:3}\includegraphics[scale=0.3]{imagesMetaTracts/crop13-a.eps}}
%  \subfloat[X Axis]{\includegraphics[width=40mm, height=40mm]{imagesMetaTracts/xaxis.eps}}\label{figur:1}\\
%  \subfloat[Y Axis]{\label{figur:2}\includegraphics[width=40mm, height=40mm]{imagesMetaTracts/yaxis.eps}}
%  \subfloat[Z Axis]{\label{figur:3}\includegraphics[width=40mm, height=40mm]{imagesMetaTracts/zaxis.eps}}
%  \caption{One of our datasets.Fig \ref{fig:orig_data}(a) provides an overview of the entire volume, the rest show slices of each of the major axis.}\label{fig:orig_data}
%\end{figure}

@@

In Fig.~\ref{fig:reliable_hessian} we see the intermediate results of the local orientation computation only at locations where $R_{H}$ is greater than the cutoff threshold. The unit vector representing the principal direction has been mapped to RGB color space. Fig.~\ref{fig:reliable_hessian}a shows the entire data set. Regions where the principal direction is parallel to the X axis are red in color, directions parallel to Y are green and those parallel to the Z are blue. Fig.~\ref{fig:reliable_hessian}b,c show 2D slices along the Z and X axes respectively. Fig.~\ref{fig:reliable_hessian}d shows a magnified region of interest. Note, the dark regions within a bundles are regions where the $R_H$ is less than the threshold and are unreliable. The bundles are not uniformly colored, the Hessians and the corresponding directions are noisy.

@@

%
\begin{figure}[tb]
\centering
%\includegraphics[width=0.45\textwidth]{images_pvis/figure3}
%\includegraphics[width=0.5\textwidth,  trim = 0mm 90mm 0mm 0mm, clip]{images_pvis/figure3}
\includegraphics[width=\linewidth]{images_pvis/reliable_hessian.pdf}
\caption{Reliable Hessians. (a) Colored according to Orientation vector mapped to RGB. (b) 2D slice along Z and (c) along X axis. (d) Magnified region marked in green square.}
\label{fig:reliable_hessian}
\end{figure}

@@

\subsection {MetaTracts generation}
\label{subsec:MetaTracts-generation}
 In this section we explain the process of MetaTracts generation in $\mathbb{R}^2$.The procedure trivially extends to $\mathbb{R}^3$. In $\mathbb{R}^2$, all the above properties hold except that cylinders are replaced by rectangles. The reliable Hessian step  generates a local orientation in $\mathbb{R}^2$ and a reliability measure $R_H$[0,1]. In Fig. \ref{fig:algo}, all regions which have unreliable Hessian have been marked in blue. Fig.~\ref{fig:algo}c shows a single cylinder of a MetaTracts. Let the seed point associated with the MetaTract be grid point $C_{p}$ (property 2 in Sec.~\ref{subsec:mtprop}). The local orientation at $C_p$ as computed in the step above is $N_P$ and is given by the dark green arrow. We generate a rectangle of length $L$ and radius $R$, the green dashed rectangle shows the rectangle generated at $C_p$ (properties 3 and 5 in Section~\ref{subsec:mtprop}). 
%
\begin{figure}[tb]
  \begin{center}
 \includegraphics[ width=\linewidth]{images_pvis/mt_gen.pdf}
  \end{center}
  \vspace{-1em}
 \caption{MetaTracts generation: (b) MetaTract generation, region with $R_H$ less than threshold. (a) close up of marked region in (b). (c) Parts of a single MetaTract.}
  \label{fig:algo}
    \vspace{-1.5em}
\end{figure}
The set of vertices which are in the green region but not overlapped by blue are possible candidates for the start point of the next cylinder in a particular MetaTract. We call these vertices \textit{``candidate vertices"}. From these we select another grid vertex which will be the start point for the next cylinder.
%\begin{figure}
%\centering
% \includegraphics[width=0.25\textwidth]{imagesMetaTracts/angle-diff.png}
%\caption{Angle between the local orientation of two cylinders that are part of the same MetaTract.}
%\label{fig:angle-diff}
%\end{figure}
Based on such assumptions on our data as local orientation and connectivity (Sec.~\ref{sec:char_data}), the order of the candidate vertices is based on the following characteristics:

\begin{itemize}[noitemsep]
\item \textit{Orientation similarity}: The orientation of the start points ($N_p$) for the consecutive cylinders should be similar. 
\item \textit{Large distance}: The MetaTracts should traverse the data using as few cylinders as possible. Thus, the distance between $C_p$ and the start point for the next cylinder should be as large as possible. We measure the distance of a candidate vertex from $C_p$ by projecting the Euclidean distance between them onto $N_p$. For example, in Fig.~\ref{fig:algo}b the distance is measured as the Euclidean distance between the purple and the orange vertices projected on to $N_p$. We refer to this perpendicular projection distance as \textit{`projected\_dist'}.

\end{itemize}
% \begin{equation}\label{eqn:algo_1}
% d=\frac{1}{3}e^{(\frac{-angle^{2}}{\alpha^{2}})}+\frac{2}{3}(1-e^{\frac{-{PerpDist}^{2}}{\beta^{2}}})
% \end{equation}
We define a priority for each of the candidate vertices based on the above factors.
For each cylinder in a MetaTract, we put its \textit{candidate vertices} in a priority queue based on Equation \ref{eqn:algo_1}.
\begin{equation}
Priority = \gamma_1 e^{(-angle^2 / \alpha^2)} + \gamma_2e^{(-projected\_dist^2 / \beta^2)}
\label{eqn:algo_1}
\end{equation}
$\gamma_1$, $\gamma_2$ are the  weights ($\mathbb{R}_{\ge 0}$)  which decide how the priority depends on the affine combination of the two factors. For all our cases, we use $\gamma_1$ = $1 / 3 $ and $\gamma_2$ = $2 / 3$. In general, we suggest $\gamma_1 +\gamma_2 = 1 $ and $\gamma_1 \leq \gamma_2$. At each iteration, we pick the top element in the priority queue to generate the corresponding cylinder, and repeat the steps. Essentially, Equation~\ref{eqn:algo_1} selects a grid point which is furthest from the current point and is going in a similar direction. This approach tackles noise/errors in local orientation better than integral curves by looking at multiple choices for vertices and avoiding intra cell interpolation in an already noisy environment. 
% 
In the example in Fig.~\ref{fig:algo}b we select the orange grid vertex next and repeat the process. We do not select the purple vertex (Fig.~\ref{fig:algo}c) because it is not a candidate vertex. If we generate tracts that have erroneous local orientations they will not be able to find further possible candidate vertices. Therefore these short tracts are removed. The MetaTracts generated are shown in Fig.~\ref{fig:length_distribution}a. The MetaTracts are colored with the mean orientation direction mapped to the RGB space. Fig.~\ref{fig:length_distribution}a shows a single slice of YZ plane. Consistent orientation is a key intrinsic feature in the data becomes visually pronounced.
We apply uniform, dense seeding to the XCT volume data to trace and generate the tracts of fiber bundles. Fig.~\ref{fig:length_distribution}b shows the MetaTracts of a particular bundle colored according to their individual length.
%
MetaTracts in a given fiber bundle do not extend the full length of the bundle. MetaTracts may have different lengths and may partially overlap.

@@

\begin{figure}[tb]
\centering
	\includegraphics[width=\linewidth]{images_pvis/figure5.pdf}
	%\includegraphics[width=0.45\textwidth,  trim = 0mm 100mm 0mm 50mm, clip]{imagesMT2014/image_length}
	\caption{(a) MetaTracts color-coded according to their mean orientations. (b) length distribution of individual MetaTracts for a particular bundle (unit for length is grid cube edge length:  $2\mu m$). A single fiber bundle is made of MetaTracts of varying length}
	\label{fig:length_distribution}
\end{figure}

@@

The output of the above step is a set of MetaTracts. The next step is to cluster the MetaTracts into bundles. 
Carbon fibers have both orientation and geometric proximity information, both of which can be used for clustering. Experimentally we found that the orientation measure was very reliable, while measures of geometric proximity had problems with fibers which partially overlap (Fig.~\ref{fig:length_distribution}b).  
Instead of creating a heuristic which artificially combines the orientation and geometric proximity measures, we first cluster based on orientation and then further subdivide each orientation cluster using geometric proximity.
We found that different clustering techniques performed preferably for different measures.  The two step clustering also made the problem more tractable and  the parameters intuitive for the end users. For the orientation clustering, we use dimension reduction followed by K-means clustering. We then use hierarchical clustering to further subdivide each cluster based on the geometric proximity measures.

@@

\subsection {Orientation based clustering}
%We use the inherent features expressed as weaving pattern of our data to pre-cluster the metaTracts into the number of classes based on the major directions of the pattern.
%It is important to note that this step broadly divides the MetaTracts into classes based on the major orientations and not individual fiber bundles. 
This step divides the individual MetaTracts into classes based on their major orientations. In order to cluster MetaTracts going in the same directions, we use a spectral embedding technique called Laplacian eigenmaps which was originally introduced by Belkin and Niyogi \cite{Belkin01}. An eigenvalue problem is solved, which maps the manifold embedded in the graph into a lower dimensional space while preserving the graph structure.
Let $G$ be the graph, we compute the eigenvalues and eigenvectors for the generalized eigenvector problem $L\textit{\textbf{f}}=\lambda D\textit{\textbf{f}}$,
%\begin{equation}\label{equn:eigenMaps}
%L\textbf{f}=\lambda D\textbf{f}
%\end{equation}
where $D$ is the diagonal weight matrix and $L$ is the Laplacian matrix. The eigenvector \textbf{${f}_{0}$} corresponding to the eigenvalue 0 is left out and the next $m$, {\textbf{${f}_{1}$} through \textbf{${f}_{m}$}} eigenvectors are used to embed in an $m$-dimensional space (see Sec.~\ref{sec:param_choices} for values of m).
 In our case, each MetaTract is a data point. We introduce a simple \textit{orientation based similarity measure}. There is no spatial information involved, just a partition of the MetaTracts based on orientation which is already inherent in the data.
Following this, given a pair of  MetaTracts, we define the edge weights between two tracts as the cosine of the maximum angle between the local orientations ($N_P$) of all pairs of start points ($C_P$) between the two MetaTracts. The edge weights give a distance matrix representing the distance between each pair of points. Using the Belkin and Niyogi algorithm we ``embed" these points in a low dimensional space where the Euclidean distance between points approximates the distance between points given by the distance matrix. 
%
%August 13 2014
%For all our tests the fibers were then embedded in a $\eta$ low dimensional space. The number of dimensions in the high dimensional space is equal to the number of datapoints or the number of MetaTracts in our case.
%For all our test cases we set $\eta$ to be five (see sec~\ref{sec:param_choices} for parameter choices for $\eta$.).
%  
We used conventional K-means for clustering this lower dimensional space. Here $K$ is supplied by domain knowledge of the number of major fiber bundle directions of the woven structure.
For our test case there are two major directions of the fiber bundles and $K$ was set to 2. Due to the advantages offered by the dimensionality reduction, even if there are curved fiber bundles the user has to provide just the major fiber bundle directions of the weaving pattern. Fig.~\ref{fig:orientation_clustering}b shows the result of the K-means clustering with the data points projected to the top three eigenvectors as the major axes. As is expected there is a clear distinction based on fiber orientation. Fig.~\ref{fig:orientation_clustering}a,c shows the MetaTracts colored according to clustering results.

>>>>

Figure~\ref{fig:data-char} provides an example of our targeted dataset. It also shows the recurring fiber bundle pattern in the final CFRP laminate, called a \textit{unit cell}.

@@

\item{ How to visualize the interaction between a particular pair of fiber bundles (weaving/braiding) or a unit cell?}

@@

 We demarcate the above queries into two parts: \textit{geometric structure} and \textit{spatial context}. Geometric structure refers to the shape, size, and orientation of a fiber bundle or a group of bundles. Spatial context refers to how two or more bundles interact with each other. For example, spatial context answers questions such as: Are these bundles in contact at a particular position in the data? What is the relative orientations of the contacting bundles?

Answering the above queries from the volume renderings of XCT datasets, or from visual inspection of particular 2D slices is non-trivial, even for experts. We detail and evaluate a method which uses visualization techniques to gain insight into our data.
We advance and interpret techniques from diffusion tensor imaging to extract and visualize geometric structures from 3D X-ray computed tomography data of the woven carbon fiber reinforced composites. The main goal of this work is to expand the state of the art in non-destructive testing through visualization of composite structures in \textit{lower resolutions} and thus \textit{larger areas} of woven fabrics.
\begin{figure}[tb]
	\centering
	%\includegraphics[width=0.45\textwidth]{imagesMT2014/algo.png}
	%\includegraphics[width=0.5\textwidth,  trim = 0mm 300mm 0mm 0mm, clip,]{imagesMT2014/image_algo}
	\includegraphics[width=\linewidth]{images_pvis/workflow.pdf}
	\caption{Flow chart of the MetaTracts approach for fiber bundle extraction}
	\label{fig:flowchart}
\end{figure}
Figure~\ref{fig:flowchart} shows an over view of our approach. Starting from the XCT data, we first generate MetaTracts, cluster the MetaTracts to create fiber bundles and then generate visualizations from the clusters. For a detailed description of the MetaTracts generation process, see Sec.~\ref{subsec:MetaTracts-generation}. For the description of plausible fiber bundle generation see Sec.~\ref{subsec:fiber-bundles}. For a discussion of the visualizations we create for our domain experts see Sec.~\ref{sec:vis}. Sec.~\ref{sec:results} describes our experimental results, Sec.~\ref{sec:param_choices} presents our parameter choices and  Sec.~\ref{sec:user_eval} summarizes our user evaluations.

@@

\subsection {Visual Analysis and Modeling of Fiber Reinforced Composites}
The approaches presented in visualization and analysis of composites mainly focus on individual objects such as fibers or pores and deals with fiber extraction from high resolution data where the individual fibers are clearly discernible. Fritz et al.~\cite{Fritz2009} proposed interactive workflows for non destructive testing practitioners to explore and quantify steel fibers in reinforced sprayed concrete. This approach allows analyzing fiber orientations based on direction transfer functions. Salaberger et al.~\cite{Salaberger2011} introduced a pipeline to extract and characterize individual fibers of fiber reinforced composites. They encode the extracted fibers as color-coded line segments in 3D and visually identifying fibers with similar orientations. Recently, Weissenbock et al.~\cite{Weissenbock2014} introduced a system for interactive exploration and analysis of fibers in fiber reinforced polymers.

Lomov et al.~\cite{Lomov2010} in a recent paper, discussed the current available solutions in geometric modeling of three dimensional composites. Modeling of the composites starts with establishing the topology of the structure, which translates to answering if a particular bundle is in contact with another at a particular position? The second builds the geometry of the model, answering questions relating to placement of bundles in space their orientations and dimensions.

%Tools based on image analysis, fiber tracking and clustering into bundles in DTI and their subsequent visualization through stream tubes and coloring schemes have received extensive research. Recently, similar techniques have been applied to fiber tracing in composite materials. 
To the best of our knowledge there is no approach focusing on direct extraction of fiber bundle structures from low resolution XCT of composite fibers, which is the main scope of this work.

@@

\begin{figure}[tb]
\centering
%\includegraphics[width=0.5\textwidth, trim = 0mm 110mm 0mm 0mm, clip,]{images_pvis/figure1}
\includegraphics[width=\linewidth]{images_pvis/data-char.pdf}
\caption{Data characteristics: (a) Rendering of the data. (b) 2D slices along the X-axis and (c) 2D slice along the Z-axis. (d) Magnified region marked in green square (c). Multiple fiber bundles cross and are indistinguishable by visual inspection alone. }
\label{fig:data-char}
\end{figure}
\section {Data Characteristic and Assumptions}
\label{sec:char_data}
Figure \ref{fig:data-char} shows dataset 1 with woven fiber bundles. Dataset 1 is 450$\times$300$\times$500 voxels in size with isotropic resolution of 2$\mu m$. The data type is unsigned 8 bit integer. Figure~\ref{fig:data-char} clearly shows the weaving structure of the composite unit cell as regional subset of the recurring fiber bundle weaving pattern. The unit cell is used as a basis structure for CFRP manufacturing. The alignment and weaving pattern of the fiber bundles in the cloth influences the strength and stiffness properties of the resulting material. Figure~\ref{fig:data-char}a shows a volume rendering of the dataset. Figure~\ref{fig:data-char}b shows 2D slices along the $X$- and Figure~\ref{fig:data-char}c in $Z$-axis. Figure~\ref{fig:data-char}d shows the magnified version of the green region of interest. 
The resolution of dataset 1 renders the individual carbon fibers in a fiber bundle hard to resolve in the 2D slice images. Figure~\ref{fig:data-char}d contains two bundles going in opposite directions. The separation between two fiber bundles is barely visible. The carbon fibers themselves do not differ much from the underlying epoxy matrix in terms of attenuation which poses an additional challenge for characterization. The fiber bundles may differ in terms of the amount of fibers in the bundle. Figure~\ref{fig:data-char}a shows the large variation in cross section sizes among the bundles.
Depending on the weaving pattern, the fiber bundles cross each other in different orientations. In general, the number of orientations is limited by the weaving pattern. Nevertheless, the weaving pattern may cause the individual bundles to be curved. In consequence individual fibers may be adjacent in Euclidean distance but belong to different bundles. 
We make the following assumptions on our data:
\begin{itemize}[noitemsep]
\item The structure embedded in the data contains fiber bundles of indiscernible fibers.
\item Local orientation: each point within a fiber bundle has a local orientation which is parallel to the corresponding area in the fiber bundle.
\item The local orientation may change gradually inside the fiber bundle.
\item The local orientation may be noisy and not reliable.
\item Connectivity: moving along the direction of a non-noisy local orientation in small increments, we will reach another neighborhood with similar local orientation.
\item Fiber bundles going in different directions only interact near the surface of contact.
\end{itemize}
%\begin{figure}[htp]
%  \centering
%    \subfloat[Overview of the dataset]{\label{figur:3}\includegraphics[scale=0.3]{imagesMetaTracts/crop13-a.eps}}
%  \subfloat[X Axis]{\includegraphics[width=40mm, height=40mm]{imagesMetaTracts/xaxis.eps}}\label{figur:1}\\
%  \subfloat[Y Axis]{\label{figur:2}\includegraphics[width=40mm, height=40mm]{imagesMetaTracts/yaxis.eps}}
%  \subfloat[Z Axis]{\label{figur:3}\includegraphics[width=40mm, height=40mm]{imagesMetaTracts/zaxis.eps}}
%  \caption{One of our datasets.Fig \ref{fig:orig_data}(a) provides an overview of the entire volume, the rest show slices of each of the major axis.}\label{fig:orig_data}
%\end{figure}

@@

In Figure~\ref{fig:reliable_hessian} we see the intermediate results of the local orientation computation only at locations where $R_{H}$ is greater than the cutoff threshold. The unit vector representing the principal direction has been mapped to RGB color space. Figure~\ref{fig:reliable_hessian}a shows the entire data set. Regions where the principal direction is parallel to the X axis are red in color, directions parallel to Y are green and those parallel to the Z are blue. Figure~\ref{fig:reliable_hessian}b,c show 2D slices along the Z and X axes respectively. Figure~\ref{fig:reliable_hessian}d shows a magnified region of interest. Note, the dark regions within a bundles are regions where the $R_H$ is less than the threshold and are unreliable. The bundles are not uniformly colored, the Hessians and the corresponding directions are noisy.

@@

%
\begin{figure}[tb]
\centering
%\includegraphics[width=0.45\textwidth]{images_pvis/figure3}
%\includegraphics[width=0.5\textwidth,  trim = 0mm 90mm 0mm 0mm, clip]{images_pvis/figure3}
\includegraphics[width=\linewidth]{images_pvis/reliable_hessian.pdf}
 \vspace{-1.5em}
\caption{Reliable Hessians. (a) Colored according to Orientation vector mapped to RGB. (b) 2D slice along Z and (c) along X axis. (d) Magnified region marked in green square.}
\label{fig:reliable_hessian}
\end{figure}

@@

\subsection {MetaTracts generation}
\label{subsec:MetaTracts-generation}
 In this section we explain the process of MetaTracts generation in $\mathbb{R}^2$.The procedure trivially extends to $\mathbb{R}^3$. In $\mathbb{R}^2$, all the above properties hold except that cylinders are replaced by rectangles. The reliable Hessian step  generates a local orientation in $\mathbb{R}^2$ and a reliability measure $R_H$[0,1]. In Figure~\ref{fig:algo}, all regions which have unreliable Hessian have been marked in blue. Figure~\ref{fig:algo}c shows a single cylinder of a MetaTracts. Let the seed point associated with the MetaTract be grid point $C_{p}$ (property 2 in Sec.~\ref{subsec:mtprop}). The local orientation at $C_p$ as computed in the step above is $N_P$ and is given by the dark green arrow. We generate a rectangle of length $L$ and radius $R$, the green dashed rectangle shows the rectangle generated at $C_p$ (properties 3 and 5 in Section~\ref{subsec:mtprop}). 
%
\begin{figure}[tb]
  \begin{center}
 \includegraphics[ width=\linewidth]{images_pvis/mt_gen.pdf}
  \end{center}
  \vspace{-1em}
 \caption{MetaTracts generation: (b) MetaTract generation, region with $R_H$ less than threshold. (a) close up of marked region in (b). (c) Parts of a single MetaTract.}
  \label{fig:algo}
    \vspace{-1.5em}
\end{figure}
The set of vertices which are in the green region but not overlapped by blue are possible candidates for the start point of the next cylinder in a particular MetaTract. We call these vertices \textit{``candidate vertices"}. From these we select another grid vertex which will be the start point for the next cylinder.
%\begin{figure}
%\centering
% \includegraphics[width=0.25\textwidth]{imagesMetaTracts/angle-diff.png}
%\caption{Angle between the local orientation of two cylinders that are part of the same MetaTract.}
%\label{fig:angle-diff}
%\end{figure}
Based on such assumptions on our data as local orientation and connectivity (Sec.~\ref{sec:char_data}), the order of the candidate vertices is based on the following characteristics:

\begin{itemize}[noitemsep]
\item \textit{Orientation similarity}: The orientation of the start points ($N_p$) for the consecutive cylinders should be similar. 
\item \textit{Large distance}: The MetaTracts should traverse the data using as few cylinders as possible. Thus, the distance between $C_p$ and the start point for the next cylinder should be as large as possible. We measure the distance of a candidate vertex from $C_p$ by projecting the Euclidean distance between them onto $N_p$. For example, in Figure~\ref{fig:algo}b the distance is measured as the Euclidean distance between the purple and the orange vertices projected on to $N_p$. We refer to this perpendicular projection distance as \textit{`projected\_dist'}.

\end{itemize}
% \begin{equation}\label{eqn:algo_1}
% d=\frac{1}{3}e^{(\frac{-angle^{2}}{\alpha^{2}})}+\frac{2}{3}(1-e^{\frac{-{PerpDist}^{2}}{\beta^{2}}})
% \end{equation}
We define a priority for each of the candidate vertices based on the above factors.
For each cylinder in a MetaTract, we put its \textit{candidate vertices} in a priority queue based on Equation \ref{eqn:algo_1}.
\begin{equation}
Priority = \gamma_1 e^{(-angle^2 / \alpha^2)} + \gamma_2e^{(-projected\_dist^2 / \beta^2)}
\label{eqn:algo_1}
\end{equation}
$\gamma_1$, $\gamma_2$ are the  weights ($\mathbb{R}_{\ge 0}$)  which decide how the priority depends on the affine combination of the two factors. For all our cases, we use $\gamma_1$ = $1 / 3 $ and $\gamma_2$ = $2 / 3$. In general, we suggest $\gamma_1 +\gamma_2 = 1 $ and $\gamma_1 \leq \gamma_2$. At each iteration, we pick the top element in the priority queue to generate the corresponding cylinder, and repeat the steps. Essentially, Equation~\ref{eqn:algo_1} selects a grid point which is furthest from the current point and is going in a similar direction. This approach tackles noise/errors in local orientation better than integral curves by looking at multiple choices for vertices and avoiding intra cell interpolation in an already noisy environment. 
% 
In the example in Figure~\ref{fig:algo}b we select the orange grid vertex next and repeat the process. We do not select the purple vertex (Figure~\ref{fig:algo}c) because it is not a candidate vertex. If we generate tracts that have erroneous local orientations they will not be able to find further possible candidate vertices. Therefore these short tracts are removed. The MetaTracts generated are shown in Figure~\ref{fig:length_distribution}a. The MetaTracts are colored with the mean orientation direction mapped to the RGB space. Figure~\ref{fig:length_distribution}a shows a single slice of YZ plane. Consistent orientation is a key intrinsic feature in the data becomes visually pronounced.
We apply uniform, dense seeding to the XCT volume data to trace and generate the tracts of fiber bundles. Figure~\ref{fig:length_distribution}b shows the MetaTracts of a particular bundle colored according to their individual length.
%
MetaTracts in a given fiber bundle do not extend the full length of the bundle. MetaTracts may have different lengths and may partially overlap.

@@

\begin{figure}[tb]
\centering
	\includegraphics[width=\linewidth]{images_pvis/figure5.pdf}
	%\includegraphics[width=0.45\textwidth,  trim = 0mm 100mm 0mm 50mm, clip]{imagesMT2014/image_length}
	\caption{(a) MetaTracts color-coded according to their mean orientations. (b) length distribution of individual MetaTracts for a particular bundle (unit for length is grid cube edge length:  $2\mu m$).}
	\label{fig:length_distribution}
\end{figure}

@@

The output of the above step is a set of MetaTracts. The next step is to cluster the MetaTracts into bundles. 
Carbon fibers have both orientation and geometric proximity information, both of which can be used for clustering. Experimentally we found that the orientation measure was very reliable, while measures of geometric proximity had problems with fibers which partially overlap (Figure~\ref{fig:length_distribution}b).  
Instead of creating a heuristic which artificially combines the orientation and geometric proximity measures, we first cluster based on orientation and then further subdivide each orientation cluster using geometric proximity.
We found that different clustering techniques performed preferably for different measures.  The two step clustering also made the problem more tractable and  the parameters intuitive for the end users. For the orientation clustering, we use dimension reduction followed by K-means clustering. We then use hierarchical clustering to further subdivide each cluster based on the geometric proximity measures.

@@

\subsection {Orientation based clustering}
%We use the inherent features expressed as weaving pattern of our data to pre-cluster the metaTracts into the number of classes based on the major directions of the pattern.
%It is important to note that this step broadly divides the MetaTracts into classes based on the major orientations and not individual fiber bundles. 
This step divides the individual MetaTracts into classes based on their major orientations. In order to cluster MetaTracts going in the same directions, we use a spectral embedding technique called Laplacian eigenmaps which was originally introduced by Belkin and Niyogi \cite{Belkin01}. An eigenvalue problem is solved, which maps the manifold embedded in the graph into a lower dimensional space while preserving the graph structure.
Let $G$ be the graph, we compute the eigenvalues and eigenvectors for the generalized eigenvector problem $L\textit{\textbf{f}}=\lambda D\textit{\textbf{f}}$,
%\begin{equation}\label{equn:eigenMaps}
%L\textbf{f}=\lambda D\textbf{f}
%\end{equation}
where $D$ is the diagonal weight matrix and $L$ is the Laplacian matrix. The eigenvector \textbf{${f}_{0}$} corresponding to the eigenvalue 0 is left out and the next $m$, {\textbf{${f}_{1}$} through \textbf{${f}_{m}$}} eigenvectors are used to embed in an $m$-dimensional space (see Sec.~\ref{sec:param_choices} for values of m).
 In our case, each MetaTract is a data point. We introduce a simple \textit{orientation based similarity measure}. There is no spatial information involved, just a partition of the MetaTracts based on orientation which is already inherent in the data.
Following this, given a pair of  MetaTracts, we define the edge weights between two tracts as the cosine of the maximum angle between the local orientations ($N_P$) of all pairs of start points ($C_P$) between the two MetaTracts. The edge weights give a distance matrix representing the distance between each pair of points. Using the Belkin and Niyogi algorithm we ``embed" these points in a low dimensional space where the Euclidean distance between points approximates the distance between points given by the distance matrix. 
%
%August 13 2014
%For all our tests the fibers were then embedded in a $\eta$ low dimensional space. The number of dimensions in the high dimensional space is equal to the number of datapoints or the number of MetaTracts in our case.
%For all our test cases we set $\eta$ to be five (see sec~\ref{sec:param_choices} for parameter choices for $\eta$.).
%  
We used conventional K-means for clustering this lower dimensional space. Here $K$ is supplied by domain knowledge of the number of major fiber bundle directions of the woven structure.
For our test case there are two major directions of the fiber bundles and $K$ was set to 2. Due to the advantages offered by the dimensionality reduction, even if there are curved fiber bundles the user has to provide just the major fiber bundle directions of the weaving pattern. Figrue.~\ref{fig:orientation_clustering}b shows the result of the K-means clustering with the data points projected to the top three eigenvectors as the major axes. As is expected there is a clear distinction based on fiber orientation. Figure~\ref{fig:orientation_clustering}a,c shows the MetaTracts colored according to clustering results.

\begin{figure}[tb] 
  \centering  	
  \includegraphics[width=\linewidth]{images_pvis/clustering.pdf}
   \vspace{-1.5em}
  %\includegraphics[width=0.45\textwidth]{imagesMT2014/image_clustering}
  	\caption{(b) Result of K means clustering with data points projected to the top three eigenvectors as major axes. (a) MetaTracts belonging to orientation cluster 1. (c) MetaTracts belonging to orientation cluster 2. MetaTracts in gray (a,c) show context.
  	(d,e) shows the result of distance based clustering on the orientation cluster. (f) shows the combined result. }
  \label{fig:orientation_clustering}
  \end{figure}
